---
title: "A 3-way crosstab table using {gt}"
author: "John Johnson"
date: "9/21/2020"
output:
  html_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
# Ignore this chunk. It silently builds the table in order to display
# it at the top of the document.

knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gt)

income.sex.age <- read_csv("data/income_by_sex_and_age.csv")
total.jobs.by.sex <- read_csv("data/total_jobs_by_sex.csv")

top.jobs <- total.jobs.by.sex %>%
  group_by(OCC2010, n) %>% 
  summarise() %>%
  arrange(desc(n)) %>%
  pull(OCC2010)

wage.tibble <- inner_join(
  # pivot wider
  income.sex.age %>%
    pivot_wider(names_from = c(age_category, SEX), values_from = median_wage),
  # pivot wider
  total.jobs.by.sex %>%
    pivot_wider(names_from = c(age_category, SEX), values_from = pct)
) %>%
  # arrange columns in the appropriate order
  select(OCC2010, starts_with("Under"), starts_with("35"), 
         starts_with("50"), `total jobs_Male` = total_Male, 
         `total jobs_Female` = total_Female, n, obs = unweighted_n) %>%
  # make OCC2010 a factor and order from most jobs to fewest
  mutate(OCC2010 = factor(OCC2010, levels = top.jobs)) %>%
  arrange(OCC2010)

wage.gap.values <- income.sex.age %>%
  pivot_wider(names_from = SEX, values_from = median_wage) %>%
  mutate(wage_gap = Male - Female)

wage.gap.table <- wage.gap.values %>%
  pivot_longer(cols = c("Male", "Female")) %>%
  select(OCC2010, age_category, name, wage_gap) %>%
  pivot_wider(names_from = c(age_category, name), values_from = wage_gap) %>%
  # arrange from most jobs to fewest
  mutate(OCC2010 = factor(OCC2010, levels = top.jobs)) %>%
  arrange(OCC2010)
purple_pal <- leaflet::colorNumeric(palette = "Purples", 
                                    domain = wage.gap.values$wage_gap)
luminance <- function(col){
  # RGB values of the color
  rgb.values <- col2rgb(col)
  
  # calculate luminance
  red.value <- rgb.values[1] * 0.299
  green.value <- rgb.values[2] * 0.587
  blue.value <- rgb.values[3] * 0.114
  
  # add the values together and divide by 255 (the max value)
  (red.value + green.value + blue.value)/255
}
luminance("orange") # for example

text_color <- function(color){
  if(luminance(color) < 0.6){
    "white"
  } else {
    "black"
  }
}

format_wage_gap <- function(gtobj, column){
  # generate color for each cell in the column
  color_value <- purple_pal(wage.gap.table %>% pull(sym(column)))
  
  # for each row in the column, perform these formatting steps
  for(i in seq_along(wage.tibble %>% pull(sym(column)))){
    gtobj <- gtobj %>%
      tab_style(style = list(
        # fill the cell background
        cell_fill(color = color_value[i]),
        # color the text black or white depending on the luminence of the background color
        cell_text(color = text_color(color_value[i]))
      ),
      locations = cells_body(columns = column, rows = i))
  }
  gtobj
}

job.gap.table <- total.jobs.by.sex %>%
  pivot_wider(names_from = SEX, values_from = pct) %>%
  mutate(job_gap = Male - Female) %>%
  select(OCC2010, `total jobs_Male` = Male, `total jobs_Female` = Female, job_gap) %>%
  mutate(`total jobs_Male` = job_gap, `total jobs_Female` = job_gap) %>%
  select(-job_gap) %>%
  mutate(OCC2010 = factor(OCC2010, levels = top.jobs)) %>%
  arrange(OCC2010)

# use the PRGn diverging palette
purple_green_pal <- leaflet::colorNumeric(palette = "PRGn", domain = job.gap.table$`total jobs_Male`,
                                          reverse = TRUE)

format_job_gap <- function(gtobj, column){
  # generate color for each cell in the column
  color_value <- purple_green_pal(job.gap.table %>% pull(sym(column)))
  
  # for each row in the column, perform these formatting steps
  for(i in seq_along(wage.tibble %>% pull(sym(column)))){
    gtobj <- gtobj %>%
      tab_style(style = list(
        # fill the cell background
        cell_fill(color = color_value[i]),
        # color the text black or white depending on the luminence of the background color
        cell_text(color = text_color(color_value[i]))
      ),
      locations = cells_body(columns = column, rows = i))
  }
  gtobj
}

final.table <- wage.tibble %>%
  gt(rowname_col = "OCC2010") %>%
  tab_spanner_delim(delim = "_") %>%
  fmt_currency(columns = c(2:7), decimals = 0, suffixing = T) %>%
  fmt_percent(columns = 8:9, decimals = 0, scale_values = F) %>%
  fmt_number(columns = 10, decimals = 1, suffixing = T) %>%
  fmt_number(columns = 11, decimals = 0) %>%
  tab_header(title = "Median annual wage for early-, mid-, and late-career workers by gender",
             subtitle = md("includes all full-time workers in the top 20 most common occupational categories from the 2014-18 [American Community Survey](https://www.census.gov/programs-surveys/acs)")) %>%
  tab_stubhead("occupation category") %>%
  tab_source_note(md("**Data source:** Steven Ruggles, Sarah Flood, Ronald Goeken, Josia Grover, Erin Meyer, Jose Pacas
                     and Matthew Sobek. *IPUMS USA: Version 10.0* [dataset]. Minneapolis, MN: IPUMS 2020. [https://doi.org/10.18128/D010.V10.0](https://doi.org/10.18128/D010.V10.0).
                     Dollars are all in 2018 values. Full-time employment is defined as at least 37 hours per typical week.")) %>%
  tab_footnote(footnote = md("IPUMS USA variable [*OCC2010*](https://usa.ipums.org/usa-action/variables/OCC2010#description_section), which is ''a harmonized occupation coding scheme based on the Census Bureau's 2010 ACS occupation classification scheme.''"),
               locations = cells_stubhead()) %>%
  tab_footnote(footnote = "The weighted estimate of total full-time workers in the United States",
               locations = cells_column_labels("n")) %>%
  tab_footnote(footnote = "The unweighted sample count",
               locations = cells_column_labels("obs")) %>%
  tab_options(heading.title.font.size = 24,
              heading.title.font.weight = "bolder",
              column_labels.font.weight = "bold") %>%
  format_wage_gap("Under 35_Male") %>%
  format_wage_gap("Under 35_Female") %>%
  format_wage_gap("35 to 50_Male") %>%
  format_wage_gap("35 to 50_Female") %>%
  format_wage_gap("50 plus_Male") %>%
  format_wage_gap("50 plus_Female") %>%
  format_job_gap("total jobs_Male") %>%
  format_job_gap("total jobs_Female") %>%
  tab_footnote(footnote = "Darker colors indicated a larger wage gap between men and women.",
               locations = cells_column_spanners(c("Under 35", "35 to 50", "50 plus"))) %>%
  tab_footnote(footnote = "Purple = more men than woman. Green = more women than men.",
               locations = cells_column_spanners("total jobs"))
```

&nbsp;
```{r table_demo, message=FALSE, echo=FALSE}
final.table
```
&nbsp;

Public opinion researchers (and others) often want to display the interaction of 3 variables in a 3-way crosstab table. The `gt` package (along with `tidyr::pivot_wider`) make this once-tedious task easy. Here is an example of how to build one of these tables using U.S. Census microdata downloaded from [IPUMS USA](https://usa.ipums.org/usa/index.shtml).


## Preparing the data
I've already summarized the individual-level microdata into two summary tables. [Click here](https://johndjohnson.info/post/processing-ipums-usa-data-for-my-income-by-gender-and-age-rstudio-table-contest-entry/) to see this process in detail.

The first table contains the median wages for workers by sex and age.
```{r}
library(tidyverse)
library(gt)

income.sex.age <- read_csv("https://raw.githubusercontent.com/jdjohn215/2020-RStudio-Table-Contest/master/data/income_by_sex_and_age.csv")
head(income.sex.age)
```

The second shows the total share of jobs by sex, as well as the total job count (weighted and unweighted).
```{r}
total.jobs.by.sex <- read_csv("https://raw.githubusercontent.com/jdjohn215/2020-RStudio-Table-Contest/master/data/total_jobs_by_sex.csv")
head(total.jobs.by.sex)
```

The rows in the table are ordered from most jobs to fewest. This correctly-ordered vector of occupation names will be useful later on.

```{r, message=FALSE}
# this is a vector of the occupation names in order of largest to smallest
top.jobs <- total.jobs.by.sex %>%
  group_by(OCC2010, n) %>% 
  summarise() %>%
  arrange(desc(n)) %>%
  pull(OCC2010)
```

The first step is widening and combining the tables into a `tibble` which will then be passed to `gt`.

```{r, message=FALSE}
wage.tibble <- inner_join(
  # pivot wider
  income.sex.age %>%
    pivot_wider(names_from = c(age_category, SEX), values_from = median_wage),
  # pivot wider
  total.jobs.by.sex %>%
    pivot_wider(names_from = c(age_category, SEX), values_from = pct)
) %>%
  # arrange columns in the appropriate order
  select(OCC2010, starts_with("Under"), starts_with("35"), 
         starts_with("50"), `total jobs_Male` = total_Male, 
         `total jobs_Female` = total_Female, n, obs = unweighted_n) %>%
  # make OCC2010 a factor and order from most jobs to fewest
  mutate(OCC2010 = factor(OCC2010, levels = top.jobs)) %>%
  arrange(OCC2010)

wage.tibble
```

## Using {gt}

Now that the `tibble` is organized, we can turn it into a `gt` object, specifying that the occupation name column should become the "stub."

```{r, message=FALSE}
wage.gt.step1 <- gt(wage.tibble, rowname_col = "OCC2010")
wage.gt.step1 %>% 
  # this is just for rmarkdown display purposes
  tab_options(container.height = px(500))
```
&nbsp;

Notice the the underscores (`_`) added to the column names when we used `pivot_wider` in step 1. These underscores indicate the column spanners. In this step, we'll also format the data presentation in each column.

```{r, message=FALSE}
wage.gt.step2 <- wage.gt.step1 %>%
  # create column spanners as already indicated in the column names by "_"
  tab_spanner_delim(delim = "_") %>%
  # format currency colums
  fmt_currency(columns = c(2:7), decimals = 0, suffixing = T) %>%
  # format percentage columns
  fmt_percent(columns = 8:9, decimals = 0, scale_values = F) %>%
  # format the weighted count as a rounded, and suffixed number
  fmt_number(columns = 10, decimals = 1, suffixing = T) %>%
  # format the actual sample count without rounding
  fmt_number(columns = 11, decimals = 0)
wage.gt.step2 %>% tab_options(container.height = px(500))
```
&nbsp;

Add explanatory text. The `gt` package makes it easy to add titles, subtitles, footnotes, and data notes. We don't need to specify footnote numbers, because `gt` assigns them automatically based on their first appearance in the table. The `md` function returns markdown formatted text.

```{r, message=FALSE}
wage.gt.step3 <- wage.gt.step2 %>%
  # add explanatory text
  tab_header(title = "Median annual wage for early-, mid-, and late-career workers by gender",
             subtitle = md("includes all full-time workers in the top 20 most common occupational categories from the 2014-18 [American Community Survey](https://www.census.gov/programs-surveys/acs)")) %>%
  tab_stubhead("occupation category") %>%
  tab_source_note(md("**Data source:** Steven Ruggles, Sarah Flood, Ronald Goeken, Josia Grover, Erin Meyer, Jose Pacas
                     and Matthew Sobek. *IPUMS USA: Version 10.0* [dataset]. Minneapolis, MN: IPUMS 2020. [https://doi.org/10.18128/D010.V10.0](https://doi.org/10.18128/D010.V10.0).
                     Dollars are all in 2018 values. Full-time employment is defined as at least 37 hours per typical week.")) %>%
  tab_footnote(footnote = md("IPUMS USA variable [*OCC2010*](https://usa.ipums.org/usa-action/variables/OCC2010#description_section), which is ''a harmonized occupation coding
                  scheme based on the Census Bureau's 2010 ACS occupation classification scheme.''"),
               locations = cells_stubhead()) %>%
  tab_footnote(footnote = "The weighted estimate of total full-time workers in the United States",
               locations = cells_column_labels("n")) %>%
  tab_footnote(footnote = "The unweighted sample count",
               locations = cells_column_labels("obs")) %>%
  tab_options(heading.title.font.size = 24,
              heading.title.font.weight = "bolder",
              column_labels.font.weight = "bold")
wage.gt.step3 %>% tab_options(container.height = px(500))
```
&nbsp;

## Formatting colors

Out of the box, `gt` provides the ability to target a column by name and rows by a logical condition for formatting. It doesn't currently have the built-in ability to simply format a cell by applying a function the value contained in that cell. In this table, I want to do some something more complicated yet.

One interesting aspect of the table is the size of the gender wage gap in each age/occupation pair. This requires the viewer to do some mental math. To help the viewer compare wage gaps across the table, I want to shade each cell by the size of the wage gap between it and its partner cell.

Doing this takes several steps. First, make a `tibble` with the same rows and columns that we want to format in our `gt` object. Except, the values in the cells of this `tibble` will be the wage gaps--the values we want to color the `gt` cells by--not the median wages. Then, write a function which uses a `for` loop to target each cell in the `gt` object individually, coloring its background based on the value in the wage gap `tibble`.

### step 1: calculate wage gaps

```{r, message=FALSE}
# first, calculate the size of the wage gap for each gender/age pair
wage.gap.values <- income.sex.age %>%
  pivot_wider(names_from = SEX, values_from = median_wage) %>%
  mutate(wage_gap = Male - Female)

# create a table with same column names and row order as the gt object
wage.gap.table <- wage.gap.values %>%
  pivot_longer(cols = c("Male", "Female")) %>%
  select(OCC2010, age_category, name, wage_gap) %>%
  pivot_wider(names_from = c(age_category, name), values_from = wage_gap) %>%
  # arrange from most jobs to fewest
  mutate(OCC2010 = factor(OCC2010, levels = top.jobs)) %>%
  arrange(OCC2010)
```

### step 2: write functions to define the cell colors

First, I make a function which returns a color name in my chosen palette for each wage gap value. I create this function using the useful `colorNumeric` function from the `leaflet` package. Because *all* of the wage gaps have the same sign (men make more than women) a sequential color palette is appropriate. I'm using the "Purples" palette from [ColorBrewer](https://colorbrewer2.org/#type=sequential&scheme=Purples&n=3).

```{r, message=FALSE}
# borrow the colorNumeric function from {leaflet}
# this new function, purple_pal, returns a color for each value
purple_pal <- leaflet::colorNumeric(palette = "Purples", 
                                    domain = wage.gap.values$wage_gap)
purple_pal(wage.gap.values$wage_gap[10]) # for example
```

Some of the background colors are quite dark, so in these cases I need to switch the font color from black to white. To do this, I create two helper functions--`luminance` calculates the brightness of each color and `text_color` chooses between "black" or "white" based on that value. My `luminance` function is based on the helpful discussion beneath [this StackOverflow question](https://stackoverflow.com/questions/596216/formula-to-determine-brightness-of-rgb-color).

```{r, message=FALSE}
luminance <- function(col){
  # RGB values of the color
  rgb.values <- col2rgb(col)
  
  # calculate luminance
  red.value <- rgb.values[1] * 0.299
  green.value <- rgb.values[2] * 0.587
  blue.value <- rgb.values[3] * 0.114
  
  # add the values together and divide by 255 (the max value)
  (red.value + green.value + blue.value)/255
}
luminance("orange") # for example

# this function returns a text color based on luminance of background color
text_color <- function(color){
  if(luminance(color) < 0.6){
    "white"
  } else {
    "black"
  }
}
text_color("orange") # for example
```

Finally, I combine all these helper functions into one function we can be applied to each relevant column in the `gt` object. Thanks to Arun Kirshna for a [helpful suggestion](https://stackoverflow.com/questions/63944953/how-to-conditionally-format-a-cell-in-a-gt-table-based-on-the-value-of-the-cel/63945239#63945239) on this step.

```{r, message=FALSE}
# create a function which formats each cell individually
format_wage_gap <- function(gtobj, column){
  # generate color for each cell in the column
  color_value <- purple_pal(wage.gap.table %>% pull(sym(column)))
  
  # for each row in the column, perform these formatting steps
  for(i in seq_along(wage.tibble %>% pull(sym(column)))){
    gtobj <- gtobj %>%
      tab_style(style = list(
        # fill the cell background
        cell_fill(color = color_value[i]),
        # color the text black or white depending on the luminence of the background color
        cell_text(color = text_color(color_value[i]))
      ),
      locations = cells_body(columns = column, rows = i))
  }
  gtobj
}
```

```{r, message=FALSE}
wage.gt.step4 <- wage.gt.step3 %>%
  format_wage_gap("Under 35_Male") %>%
  format_wage_gap("Under 35_Female") %>%
  format_wage_gap("35 to 50_Male") %>%
  format_wage_gap("35 to 50_Female") %>%
  format_wage_gap("50 plus_Male") %>%
  format_wage_gap("50 plus_Female")
wage.gt.step4 %>% tab_options(container.height = px(500))
```

### step 3: repeat for job gaps

I also want to color the cell backgrounds of the job gap columns, but this time a diverging color palette is appropriate because some professions have more women than men and others have more men than women. I use another ColorBrewer palette, "[PRGn](https://colorbrewer2.org/#type=diverging&scheme=PRGn&n=3)," because it is colorblind-safe. 

```{r, message=FALSE}
# build the tibble of job gap values
job.gap.table <- total.jobs.by.sex %>%
  pivot_wider(names_from = SEX, values_from = pct) %>%
  mutate(job_gap = Male - Female) %>%
  select(OCC2010, `total jobs_Male` = Male, `total jobs_Female` = Female, job_gap) %>%
  mutate(`total jobs_Male` = job_gap, `total jobs_Female` = job_gap) %>%
  select(-job_gap) %>%
  mutate(OCC2010 = factor(OCC2010, levels = top.jobs)) %>%
  arrange(OCC2010)

# use the PRGn diverging palette
purple_green_pal <- leaflet::colorNumeric(palette = "PRGn", domain = job.gap.table$`total jobs_Male`,
                                          reverse = TRUE)

format_job_gap <- function(gtobj, column){
  # generate color for each cell in the column
  color_value <- purple_green_pal(job.gap.table %>% pull(sym(column)))
  
  # for each row in the column, perform these formatting steps
  for(i in seq_along(wage.tibble %>% pull(sym(column)))){
    gtobj <- gtobj %>%
      tab_style(style = list(
        # fill the cell background
        cell_fill(color = color_value[i]),
        # color the text black or white depending on the luminence of the background color
        cell_text(color = text_color(color_value[i]))
      ),
      locations = cells_body(columns = column, rows = i))
  }
  gtobj
}

wage.gt.step5 <- wage.gt.step4 %>%
  format_job_gap("total jobs_Male") %>%
  format_job_gap("total jobs_Female")
wage.gt.step5 %>% tab_options(container.height = px(500))
```

## Final output

As a final step, I add some additional footnotes explaining the cell background colors. Notice how `gt` automatically renumbers all the footnotes appropriately.

```{r, message=FALSE}
# add footnotes explaining the coloring of the cells
wage.gt.step6 <- wage.gt.step5 %>%
  tab_footnote(footnote = "Darker colors indicated a larger wage gap between men and women.",
               locations = cells_column_spanners(c("Under 35", "35 to 50", "50 plus"))) %>%
  tab_footnote(footnote = "Purple = more men than woman. Green = more women than men.",
               locations = cells_column_spanners("total jobs"))
wage.gt.step6 %>% tab_options(container.height = px(500))
```

### saving the table

The function `gtsave` function offers several ways to save your table, including as a `png` or `pdf`. In this example, I save the output as an `html` file.

```
gtsave(wage.gt.step4, "WageAgeGenderTable.html", inline_css = TRUE,
       path = "/your/file/directory/")
```
